Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-09-16 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Make several minor improvements in language and proofs

\begin{abstract}
This paper presents detailed formalizations of ontological arguments in a simple modal natural deduction calculus. The first formal proof closely follows the hints in Scott's manuscript about G\"odel's argument and fills in the gaps, thus verifying its correctness. The second formal proof improves the first one, by relying on the weaker modal logic {\bf KB} instead of the stronger modal logic {\bf S5}; and by avoiding the introduction of the equality predicate. The second proof is also technically shorter than the first one, because it eliminates unnecessary detours and uses Axiom 1 for the positivity of properties only once. The third and fourth proofs formalize, respectively, Anderson's and Bj{\o}rdal's variants of the ontological argument, which are known to be immune to modal collapse.
\end{abstract}

@@

G\"odel built on Leibniz's work \citep{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \citep{Goedel} as well as in notes by Scott \citep{scott}, in whom he confided his proof.

@@

The aim of this paper is to present detailed formalizations of variants of ontological arguments in a natural deduction calculus. For a comprehensive introduction to natural deduction, the reader can consult \citep{prawitz} or \citep{gabbay}. The natural deduction calculus proposed and used in this paper is an extension of a standard natural deduction calculus for (non-modal) higher-order logic \citep{BenzmuellerBrownKohlhase} with new introduction and elimination rules for modal operators (as presented in Section \ref{sec:calculus}). This calculus is sound and complete relative to the non-modal calculus extended with Axiom K and the Necessitation rule (as demonstrated in Subsection \ref{sec:compl-sound}). There are various higher-order logics \citep{BenzmuellerBrownKohlhase} and various higher-order modal logics \citep{Muskens}. The calculus proposed here can be regarded as a minimalistic calculus for a rigid higher-order modal logic K without extensionality principles. With additional axioms for stronger modal logics, such as KB and S5, this calculus is sufficient to achieve the stated aim.

The natural deduction style was chosen for four reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of proof calculi (e.g. Fitting's tableaux \citep{fitting} or Sobel's sentential modal calculus \citep{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \citep{coq} and Isabelle \citep{isabelle}, natural deduction formalizations can be verified step-by-step in such proof assistants, and we have in fact done this \citep{CSRModalLogicInCoq}. And finally, in comparison to other kinds of proof systems (e.g. resolution, tableaux, sequent calculi), natural deduction can be considered less technical, more human-readable and more intuitive, because it relies on natural principles of reasoning such as modus ponens (implication elimination) and reasoning by assumption (implication introduction); these qualities are conducive to rendering the formalizations of ontological arguments accessible to a wider audience.

The first main contribution of this paper is a detailed formalization of Scott's version \citep{scott} of G\"odel's ontological argument \citep{Goedel} (as shown in Section \ref{sec:scott}) in the proposed natural deduction calculus. The second main contribution of the paper is a new proof (shown in Section \ref{sec:newproof}), also in natural deduction style. In contrast to Scott's proof \citep{scott}, which requires the modal logic \textbf{S5}, the new proof requires only the weaker modal logic \textbf{KB}. The new proof also does not rely on the equality predicate and is much shorter. Although there have been claims (as discussed in more detail in Section \ref{sec:newproof}) that \textbf{KB} is sufficent and the equality predicate is superfluous, formal proofs substantiating these claims were missing up to now.

A major criticism against G\"odel's argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. This is discussed in greater detail in Section \ref{sec:collapse}, where a natural deduction derivation of the collapse is presented. Many recent works on the ontological argument have proposed modifications of the argument that do not entail modal collapse. In this paper we present natural deduction formalizations of two of these solutions: Anderson's emendation (in Section \ref{sec:anderson}) and Bj{\o}rdal's variant (in Section \ref{sec:bjordal}).

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek, on the other hand were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
relesed, respectively, in Isabellle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled it to be
verified with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a so far failed attempt to formalize Avicenna's
argument in PVS as well.

@@

In every inference rule, (sub-)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas changes due to a substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub-)formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

Applications of the $\beta$ rule are silently ommitted. Double lines are used to abbreviate tedious propositional reasoning steps in the derivations where the reasoning is otherwise clear. Dashed lines are used to refer to an axiom or theorem with the proof shown elsewhere. When proof trees are too large to fit on the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

The calculus having only the rules shown in Figures \ref{fig:PropositionalRules}, \ref{fig:Classical} and \ref{fig:QuantifierRules} is named \ND. This calculus is essentially the same as the simplest natural deduction calculus in the family of calculi discussed in \citep{BenzmuellerBrownKohlhase}. The main difference is that they take disjunction as a primitive connective instead of conjunction, and they use a sequent notation, where the assumptions are listed in the left side of the sequent. The calculus with the new rules for modal operators shown in Figure \ref{fig:NDK} is named \NDK.

@@

\begin{theorem}
\label{theorem:completeness}
{\NDK} is complete, relative to {\ND} extended with axiom K and the necessitation rule.
\end{theorem}
\begin{proof}
The necessitation rule can be immediately simulated with the $\nec_I$ rule. Axiom K can be derived in {\NDK} as shown below:

@@

\begin{theorem}
\label{theorem:soundness}
{\NDK} is sound, relative to {\ND} extended with axiom K and the necessitation rule.
\end{theorem}
\begin{proof}
It is necessary to show that {\NDK} proofs of the following form can be translated to proofs in {\ND} extended with the axiom K and the necessitation rule.

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \citep{hajek}). Nevertheless, some authors of variants \citep[footnotes 11 and 14]{and90} of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

\section{Some Useful Derivable Modal Principles}
\label{sec:ModalLogics}

From an axiomatic point of view, modal logics differ with respect to which additional axiom schemas they admit. Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accessibility relation in Kripke models (See for example \citep{fitting2} or \citep[Section 3.1]{blackburn} on Frame definability): T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, K and B are satisfied; and in the models of logic {\bf B}, K, T and B are satisfied.

@@

\begin{lemma*}[Reflexivity]
$$
s=s
$$
\end{lemma*}

@@

\begin{footnotesize}
\begin{prooftree}
\AXC{}\RightLabel{$1$}
\UIC{$P s$}\RightLabel{$\imp_I^1$}
\UIC{$P s\imp P s$}
\AXC{}\RightLabel{$2$}
\UIC{$P s$}\RightLabel{$\imp_I^2$}
\UIC{$P s\imp P s$}\RightLabel{$\wedge_I$}
\BIC{$P s\biimp P s$}\RightLabel{$\forall_I$}
\UIC{$\forall P. P s\biimp P s$}\dottedLine
\UIC{$s=s$}
\end{prooftree}
\end{footnotesize}

@@

In this section we present a detailed formalization of Scott's proof in the natural deduction calculus defined in Section \ref{sec:calculus}. The reasoning in Scott's manuscript has been reproduced step-by-step and all reasoning gaps are filled in by using the deduction considered most natural. All derived formulas that do appear in Scott's manuscript are marked with a $\dagger$ here. Unmarked formulas were derived in the process of filling the gaps between the marked formulas mentioned in Scott's manuscript.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
		\AXC{Reflexivity} \dashedLine
		\UIC{$\omega: \gamma = \gamma$} \RightLabel{$\imp_I$}
		\UIC{$\omega: \rho(\gamma) \imp \gamma = \gamma$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. (\rho(x) \imp x = x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec \all x. (\rho(x) \imp x = x)\scott $} \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x = x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x = x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x = x) \imp P(\lambda x. x = x)$} \RightLabel{$\imp_E$}
		\BIC{$[\subproof_1] \qquad P(\lambda x. x = x)\scott$} 
\end{prooftree}
\end{small}

@@

Comparing this new proof of Theorem 1 with the proof of Theorem 1 indicated by G\"odel and Scott, it is noticeable that new proof is significantly shorter. The key idea to obtain a shorter proof is to instantiate the second universally quantified property of Axiom 2 by $\lambda x. \neg \rho(x)$ instead of using the predicate symbols $=$ or $\neq$ as G\"odel and Scott did. This eliminates the reliance of the ontological argument on the presence of $=$ and $\neq$ in the logical language. Furthermore, Axiom 2 is then needed only once.

@@

\begin{corollary}
\label{N:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof} The proof of this corollary is identical to the proof of the same corollary shown in Section \ref{sec:scott}.
% \begin{prooftree}
% \AXC{Axiom \ref{A3}} \dashedLine
% \UIC{$P(G)$}
% \AXC{Theorem \ref{T1} for G} \dashedLine
% \UIC{$ P(G) \imp \pos \ex x.G(x) $} \RightLabel{$\imp_E$}
% \BIC{$\pos \ex x. G(x)$}
% \end{prooftree}
\end{proof}

@@

This phenomenon is very intriguing: the new proof of Lemma 1 is technically simpler (because it has fewer inferences), but Scott's proof of Lemma 1 with Theorem 2 can be considered intuitively simpler (because it is arguably easier to understand). This constitutes an interesting case for Hilbert's 24th Problem \citep{hilbert}, which asks for criteria to properly compare the simplicity of proofs. While Hilbert had mathematical proofs in mind, insights into the 24th problem could be gained by analyzing philosophical proofs as well. As G\"odel suggested, ``a scientific (exact) philosophy and theology [\ldots] is also most highly fruitful for science''.

@@

Curiously, in the new proof of Theorem 3, the actual existence of a God-like being ($\ex x. G(x)$) is proven \emph{before} its necessity ($\nec \ex x. G(x)$). This contrasts with G\"odel's proof, where the actual existence is never actually proven. It could, nevertheless, be derived as a trivial corollary by applying the modal axiom T.

>>>>

\begin{abstract}
This paper presents detailed formalizations of ontological arguments in a simple modal natural deduction calculus. The first formal proof closely follows the hints in Scott's manuscript about G\"odel's argument and fills in the gaps, thus verifying its correctness. The second formal proof improves the first one, by relying on the weaker modal logic {\bf KB} instead of the stronger modal logic {\bf S5}; and by avoiding the equality relation. The second proof is also technically shorter than the first one, because it eliminates unnecessary detours and uses Axiom 1 for the positivity of properties only once. The third and fourth proofs formalize, respectively, Anderson's and Bj{\o}rdal's variants of the ontological argument, which are known to be immune to modal collapse.
\end{abstract}

@@

G\"odel built on Leibniz's work \citep{adams} and brought the ontological argument to a modern form using a modal logic with higher-order quantification over properties. In this setting, he gave precise axioms describing the notion of \emph{positive} property and defined God as a being that has all positive properties. G\"odel's work was saved in his own notes \citep{Goedel} as well as in notes by Scott \citep{scott}, to whom he confided his proof.

@@

The aim of this paper is to present detailed formalizations of variants of ontological arguments in a natural deduction calculus. For a comprehensive introduction to natural deduction, the reader can consult \citep{prawitz} or \citep{gabbay}. The natural deduction calculus proposed and used in this paper is an extension of a standard natural deduction calculus for (non-modal) higher-order logic \citep{BenzmuellerBrownKohlhase} with new introduction and elimination rules for modal operators (as presented in Section \ref{sec:calculus}). This calculus is sound and complete relative to the non-modal calculus extended with Axiom K and the Necessitation rule (as demonstrated in Subsection \ref{sec:compl-sound}). There are many higher-order logics \citep{BenzmuellerBrownKohlhase} and higher-order modal logics \citep{Muskens}. The calculus proposed here can be regarded as a minimalistic calculus for a rigid higher-order modal logic K without extensionality principles. With additional axioms for stronger modal logics, such as KB and S5, this calculus is sufficient to achieve the stated aim.

The natural deduction style was chosen for four reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of proof calculi (e.g. Fitting's tableaux \citep{fitting} or Sobel's sentential modal calculus \citep{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with the natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \citep{coq} and Isabelle \citep{isabelle}, natural deduction formalizations can be verified step-by-step in such proof assistants, and we have in fact done this \citep{CSRModalLogicInCoq}. And finally, in comparison to other kinds of proof systems (e.g. resolution, tableaux, sequent calculi), natural deduction can be considered less technical, more human-readable and more intuitive, because it relies on natural principles of reasoning such as modus ponens (implication elimination) and reasoning by assumption (implication introduction); these qualities are conducive to rendering the formalizations of ontological arguments accessible to a wider audience.

The first main contribution of this paper is a detailed formalization of Scott's version \citep{scott} of G\"odel's ontological argument \citep{Goedel} (as shown in Section \ref{sec:scott}) in the proposed natural deduction calculus. The second main contribution of the paper is a new proof (shown in Section \ref{sec:newproof}), also in natural deduction style. In contrast to Scott's proof \citep{scott}, which requires the modal logic \textbf{S5}, the new proof requires only the weaker modal logic \textbf{KB}. The new proof also does not rely on the equality relation and is much shorter. Although there have been claims (as discussed in more detail in Section \ref{sec:newproof}) about the sufficiency of \textbf{KB} and the superfluousness of the equality relation, formal proofs substantiating these claims were missing up to now.

A major criticism against G\"odel's argument is an undesirable consequence of the stipulated axioms called \emph{modal collapse}. This is discussed in greater detail in Section \ref{sec:collapse}, where a natural deduction derivation of the collapse is presented. Many recent works on the ontological argument have proposed modifications of the argument that do not entail modal collapse. In this paper we present natural deduction formalizations of two of these solutions: Anderson's emendation (in Section \ref{sec:anderson}) and Bj{\o}rdal's variant (in Section \ref{sec:bjordal}).

\subsection{Related Work}

An early unpublished draft of Section \ref{sec:newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{controversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
released, respectively, in Isabellle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled its
verification with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a failed attempt to formalize Avicenna's
argument in PVS as well.

@@

In every inference rule, (sub-)formulas in the conclusion correspond\footnote{This correspondence could be defined formally, but it would be tedious to do so.} to formulas in the premises (even in the case of quantifier rules, where the formulas differ due to substitution). For any expression occurring both in a formula in the premise and in its corresponding (sub-)formula in the conclusion, the occurrence in the premise is said to be an \emph{ancestor} of the occurrence in the conclusion, and the occurrence in the conclusion is said to be a \emph{descendant} of the occurrence in the premise.

@@

Applications of the $\beta$ rule are silently ommitted. Double lines are used to abbreviate propositional reasoning steps when they are tedious and too space-consuming. Dashed lines are used to refer to an axiom or theorem with the proof shown elsewhere. When proof trees are too large to fit on the page, some branches may be displayed separately. In such cases, the conclusion of the branch and the location in the main proof tree where the branch belongs are annotated with the same symbol (a subscripted $\subproof$). Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

The calculus having only the rules shown in Figures \ref{fig:PropositionalRules}, \ref{fig:Classical}, \ref{fig:QuantifierRules} and \ref{fig:ReductionRules} is named \ND. This calculus is essentially the same as the simplest natural deduction calculus in the family of calculi discussed in \citep{BenzmuellerBrownKohlhase}. The main difference is that they take disjunction as a primitive connective instead of conjunction, and they use a sequent notation, where the assumptions are listed in the left side of the sequent. The calculus with the new rules for modal operators shown in Figure \ref{fig:NDK} is named \NDK.

@@

\begin{theorem}
\label{theorem:completeness}
{\NDK} is complete relative to {\ND} extended with axiom K and necessitation.
\end{theorem}
\begin{proof}
The necessitation rule can be immediately simulated with the $\nec_I$ rule. Axiom K can be derived in {\NDK} as shown below:

@@

\begin{theorem}
\label{theorem:soundness}
{\NDK} is sound relative to {\ND} extended with axiom K and necessitation.
\end{theorem}
\begin{proof}
It is necessary to show that {\NDK} proofs of the following form can be translated to proofs in {\ND} extended with the axiom K and the necessitation rule.

@@

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible world semantics}, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is tacitly assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy for the ontological argument has already been contested \citep{fitting}. Another assumption made here is that the quantification domains are constant (i.e. independent of the possible worlds). Neither G\"odel's manuscript nor Scott's manuscript reveal whether they use constant or varying domains, and this is also the case for many variants (e.g. \citep{hajek}). Nevertheless, some authors of variants of G\"odel's ontological argument do explicitly state a preference for varying domains. Our choice of constant domains is motivated by simplicity \citep[footnotes 11 and 14]{and90}. As is well-known, varying domain quantifiers (also known as \emph{actualistic} quantifiers) can be simulated by constant domain quantifiers (also known as \emph{possibilistic} quantifiers) guarded by a primitive \emph{existence} predicate\footnote{Formally, $\forall^e x. F[x] \equiv \forall x. E(x) \imp F[x]$ and $\exists^e x. F[x] \equiv \exists x. E(x) \wedge F[x]$.}; therefore, the formalization of ontological arguments with varying domain quantification would be feasible in the natural deduction calculus presented here as well.

\section{Some Useful Derivable Principles}
\label{sec:ModalLogics}

From an axiomatic point of view, modal logics stronger than {\bf K} differ from each other with respect to which axiom schemas they admit in addition to K (and necessitation). Some common axiom schemas of relevance to the ontological argument are T ($\nec A \imp A$), B ($A \imp \nec \pos A$), 4 ($\nec A \imp \nec \nec A $) and 5 ($\pos A \imp \nec \pos A$). From a semantic point of view, these axioms correspond to geometric frame conditions that must be satisfied by the accessibility relation in Kripke models (See for example \citep{fitting2} or \citep[Section 3.1]{blackburn} on Frame definability): T corresponds to reflexivity, B corresponds to symmetry, 4 corresponds to transitivity and 5 corresponds to euclidianity. In the models of logic {\bf S5}, all these axioms are satisfied; in the models of logic {\bf KB}, B is satisfied; and in the models of logic {\bf B}, T and B are satisfied.

@@

The \emph{reflexivity} of equality is a theorem in higher-order logics where equality is defined as Leibniz's equality. A natural deduction proof can be easily constructed, as shown below.

\begin{lemma*}[Reflexivity]
$$
\forall x. x = x
$$
\end{lemma*}

@@

\begin{footnotesize}
\begin{prooftree}
\AXC{}\RightLabel{$1$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^1$}
\UIC{$P(\alpha) \imp P(\alpha)$}
\AXC{}\RightLabel{$2$}
\UIC{$P(\alpha)$}\RightLabel{$\imp_I^2$}
\UIC{$P(\alpha) \imp P(\alpha)$}\RightLabel{$\wedge_I$}
\UIC{$(P(\alpha) \imp P(\alpha)) \wedge (P(\alpha) \imp P(\alpha))$} \dottedLine
\BIC{$P(\alpha) \biimp P(\alpha)$}\RightLabel{$\forall_I$}
\UIC{$\forall P. P(\alpha) \biimp P(\alpha)$}\dottedLine
\UIC{$\alpha=\alpha$} \RightLabel{$\forall_I$}
\UIC{$\forall x. x=x$} 
\end{prooftree}
\end{footnotesize}

@@

In this section we present a detailed formalization of Scott's proof in the natural deduction calculus defined in Section \ref{sec:calculus}. The reasoning in Scott's manuscript has been reproduced step-by-step and reasoning gaps are completed by using the deduction steps considered most natural. All derived formulas that do appear in Scott's manuscript are marked with a $\dagger$ here. Unmarked formulas were derived in the process of filling the gaps between the marked formulas.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$P(\rho)$}
		\AXC{Reflexivity} \dashedLine
    \UIC{$\omega: \forall x. x = x$} \RightLabel{$\forall_E$}
		\UIC{$\omega: \gamma = \gamma$} \RightLabel{$\imp_I$}
		\UIC{$\omega: \rho(\gamma) \imp \gamma = \gamma$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. (\rho(x) \imp x = x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec \all x. (\rho(x) \imp x = x)\scott $} \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x = x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x = x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec \all x. (\rho(x) \imp x = x) \imp P(\lambda x. x = x)$} \RightLabel{$\imp_E$}
		\BIC{$[\subproof_1] \qquad P(\lambda x. x = x)\scott$} 
\end{prooftree}
\end{small}

@@

Comparing this new proof of Theorem 1 with the proof of Theorem 1 indicated by G\"odel and Scott, it is noticeable that the new proof is significantly shorter. The key idea to obtain a shorter proof is to instantiate the second universally quantified property of Axiom 2 by $\lambda x. \neg \rho(x)$ instead of using the relation symbols $=$ or $\neq$ as G\"odel and Scott did. This eliminates the reliance of the ontological argument on the presence of $=$ and $\neq$ in the logical language. Furthermore, Axiom 2 is then needed only once.

@@

\begin{corollary}
\label{N:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof} The proof of this corollary is identical to the proof shown in Section \ref{sec:scott}.
% \begin{prooftree}
% \AXC{Axiom \ref{A3}} \dashedLine
% \UIC{$P(G)$}
% \AXC{Theorem \ref{T1} for G} \dashedLine
% \UIC{$ P(G) \imp \pos \ex x.G(x) $} \RightLabel{$\imp_E$}
% \BIC{$\pos \ex x. G(x)$}
% \end{prooftree}
\end{proof}

@@

This phenomenon is intriguing: the new proof of Lemma 1 is technically simpler (because it has fewer inferences), but Scott's proof of Lemma 1 with Theorem 2 can be considered intuitively simpler (because it is easier to understand). This constitutes an interesting case for Hilbert's 24th Problem \citep{hilbert}, which asks for criteria to properly compare the simplicity of proofs. While Hilbert had mathematical proofs in mind, insights into the 24th problem could be gained by analyzing philosophical proofs as well. As G\"odel suggested, ``a scientific (exact) philosophy and theology [\ldots] is also most highly fruitful for science''.

@@

Curiously, in the new proof of Theorem 3, the actual existence of a God-like being ($\ex x. G(x)$) is proven \emph{before} its necessity ($\nec \ex x. G(x)$). This contrasts with G\"odel's proof, where the actual existence is never actually proven (although it could, nevertheless, be derived as a trivial corollary by applying the modal axiom T).
